### 2018-10-16 manuscript.rmd
more outline and stubs

There has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws].

Defining Porifera-sister and Ctenophora-sister... XXX Explain that it is about support for calde of all animals except Ctenophora or Porifera...

@@

Models of molecular evolution have several components that each consider different aspects of the evolutionary process. The models that have been used to model protein evolution in studies of the animal root have largely differed according to three components: the exchangability matrix $E$, the rate of evolution, and the state equilibrium frequencies $\Pi$.

The Exchangeability matrix $E$ describes the rate at which one amino acid changes to another. Exchangability matrices have been used in the studies under consideration here include:

@@

- WAG [@Whelan:2001ds] is an emprically derived exchangability matrix based on a dataset of 182 globular protein families. It has no free parameters to estimate since all off-diagonal elements are set according to values estimated from this particular sample dataset.

- LG [@Le:2008fp], like WAG, is an emprically derived exchangability matrix. It is based on a much larger set of genes, and variation in rates across sites was taken into consideration when it was calculated. It has no free parameters to estimate since all off-diagonal elements are set according to values estimated from this particular sample dataset.

- GTR, the General Time Reversible exchangability matrix, has free parameters for all off-diagonal elements that describe the exchangability of different amino acids. It is consrained so that changes are reversible, *i.e.* the rates above the diagonal are the same as those below the diagonal. This leaves 190 parameters that must be estimated from the data long with the other model parameters and the phylogenetic tree topology. This estimation requires a considerable amount of data and computational power, but if successful has the advantage of being based on the dataset at hand rather than a different dataset (as for LG and WAG).

While the exchancheability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled. There are couple approaches that have been used in the studies considered here:

- Site homogenous rates. The rates of evolution are assumed to be the same at all sites in the amino acid alignment.

@@

- Empirical site homogenous. The frequency of each amino acid is observed from the matrix under consideration and applied to homogenously to all sites in the matrix.

- Estimated site homogenous. The frequency of each amino acid is inferred along with other model parameters, under the assumption that it is the same at all sites.

@@

Models can be assembled by selecting different options for all these different components. The models that are applied in practice area heavily influenced by engineering and computational costs, as well as convention. For example, on the questions considered here F81 and GTR exchangability matrices have only been used in combination with CAT site heterogeneous models of equilibrium frequency. LG and WAG exchangability matrices have only been used with site homogenous estimates of equilibrium frequency. This is further confused by the abreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to variation across sites and WAG a particular exchangability matrix. CAT is generally shorthand for F81+CAT and WAG is shorthand for WAG+homogenous equilibrium frequency estimation. One could, though, run a WAG+CAT model.

To avoid confusion on this point, we always specify the exchangability matrix first, followed by modifiers that describe accomodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT) or rate (*e.g.*, Gamma). If there are no modifiers, then it is implied that site homogenous models are used.

@@

@@

Sensitivity to ingroup sampling has recieved less attention than sensitivity to outgroup sampling. This may be because results have tended to be more sensitive to outgroup sampling.

@@

@@

@@

One of the challenges of interpretting support for the placement of the animal root across studies is that different programs, sofware versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways. Here we reanalyze the primary matrices from each study under consistent conditions with iqtree. We selected this tool because it has greater model flexibility than other tools and is very fast.

We

@@

@@

@@

Author contributions

@@

## Supplemental Information

@@

The data matrix was constructed using a semi-automated approach. Genes were translated into proteins, promiscuous domains were masked, all gene sequences from all species were compared to each other with blastp, genes were clustered based on this similarity with TribeMCL [@Enright:2002uq], and these clusters were filtered to remove those with poor taxon sampling and high rates of lineage-specific duplciations. Gene trees were then constructed, and in clades of sequences all from the same species all but one sequence were removed (these groups are often due to assembly errors). The remaining gene trees with more than one sequence for any taxon were then manually inspected. If strongly supported deep nodes indicative of paralogy were found, the entire gene was discarded. If the duplications for a a small number of taxa were unresolved, all genes from those taxa were excluded. Genes were then realigned and sites were filtered with Gblocks [@Castresana:2000vy], resulting in a 77 taxon matrix. Some taxa in this matrix were quite unstable, which obscured other strongly-supported relationships. Unstable taxa were identified with leaf stability indeces [@Thorley:1999kg], as implemented in phyutility [@Smith:2008gb], and removed from the matrix. This resulted in the 64-taxon matrix that is the focus of most of their analyses. Phylogenetic analyses were conducted under the F81+CAT model in phylobayes [@Lartillot:2004dq], and under the WAG model in MrBayes [@Ronquist:2003hx] and RAxML [@Stamatakis:2006wc].

@@

Pick *et al.* [@Pick:2010eb] sought to test wether Ctenophora-sister was an artefact of insufficient taxon sampling. They added new and additional published sequence data to the 64-taxon matrix of Dunn *et al.* [@Dunn:2008ky]. The new taxa included 12 sponges, 1 ctenophore, 5 cnidarians, and *Trichoplax*. They firther modified the matrix by removing 2,150 sites that were poorly sampled or aligned. They considered two different sets of outgroups: Choanoflagellata (resulting in Choanimalia) and the same sampling as Dunn *et al.* (resulting in Opisthokonta).

All their analyses were conducted under the F81+CAT+Gamma model in phylobayes [@Lartillot:2004dq], in both a bayesian framework and with bootstrapping. All analyses have the same ingroup sampling and site removal so it isn't possible to independently assess the impact of these factors. Analyses with Choanimalia sampling recovered Porifera-sister with 72% posterior probability (PP) and 91% bootstrap support (BS). With broader Opisthokonta sampling, support for Porifera-sister is 84% PP. This is an inveresting case where increased outgroup sampling leads to increased support for Porifera-sister.

The authors argue that previous results supporting Ctenophora-sister "are artifacts stemming from insufficient taxon sampling and long-branch attraction (LBA)" and that "this hypothesis should be rejected". Thout the posterior probabilities supporting Porifera-sister are not strong, they conclude:

@@

They also investigated saturation, and coclude that Dunn *et al.* [@Dunn:2008ky] is more saturated than Philippe *et al.* 2009 [Philippe:2009hh]. Note that the Pick *et al.* [@Pick:2010eb] dataset is not reanalyzed here because partition data are not available, and due to site filtering the partition file from Dunn *et al.* [@Dunn:2008ky] cannot be applied to this matrix.

>>>>

In the past decade there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister and Porifera-sister (Fig XXOverview) emerging as the two primary hypotheses. Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested. In contrast to the lack of debate about the position of Porifera, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative. Since then many more studies have been published, some supporting Ctenophora-sister, some Porifera-sister, and some neither. As it has become clear that this is a very difficult phylogenetic challenge, and the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem. Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support. Here we synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, and reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies. We hope that this provides an integrative overview of the challenge and provides direction for future studies. We also hope that the work we have done here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop methods to address difficult phylogenetic problems.

@@

Models of molecular evolution have several components that each consider different aspects of the evolutionary process. The models that have been used to model protein evolution in studies of the animal root have largely differed according to three components: the exchangeability matrix $E$, the rate of evolution, and the state equilibrium frequencies $\Pi$.

The exchangeability matrix $E$ describes the rate at which one amino acid changes to another. Exchangeability matrices have been used in the studies under consideration here include:

@@

- WAG [@Whelan:2001ds] is an empirically derived exchangeability matrix based on a dataset of 182 globular protein families. It has no free parameters to estimate since all off-diagonal elements are set according to values estimated from this particular sample dataset.

- LG [@Le:2008fp], like WAG, is an empirically derived exchangeability matrix. It is based on a much larger set of genes, and variation in rates across sites was taken into consideration when it was calculated. It has no free parameters to estimate since all off-diagonal elements are set according to values estimated from this particular sample dataset.

- GTR, the General Time Reversible exchangeability matrix, has free parameters for all off-diagonal elements that describe the exchangeability of different amino acids. It is constrained so that changes are reversible, *i.e.* the rates above the diagonal are the same as those below the diagonal. This leaves 190 parameters that must be estimated from the data long with the other model parameters and the phylogenetic tree topology. This estimation requires a considerable amount of data and computational power, but if successful has the advantage of being based on the dataset at hand rather than a different dataset (as for LG and WAG).

While the exchangeability matrix describes the relative rate of different changes between amino acids, the actual rate can be further scaled. There are couple approaches that have been used in the studies considered here:

- Site homogeneous rates. The rates of evolution are assumed to be the same at all sites in the amino acid alignment.

@@

- Empirical site homogeneous. The frequency of each amino acid is observed from the matrix under consideration and applied to homogeneously to all sites in the matrix.

- Estimated site homogeneous. The frequency of each amino acid is inferred along with other model parameters, under the assumption that it is the same at all sites.

@@

Models can be assembled by selecting different options for all these different components. The models that are applied in practice area heavily influenced by engineering and computational costs, as well as convention. For example, on the questions considered here F81 and GTR exchangeability matrices have only been used in combination with CAT site heterogeneous models of equilibrium frequency. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to variation across sites and WAG a particular exchangeability matrix. CAT is generally shorthand for F81+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. One could, though, run a WAG+CAT model.

To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT) or rate (*e.g.*, Gamma). If there are no modifiers, then it is implied that site homogeneous models are used.

@@

XXX

**Fig XXOutgroup.** The animals and their outgroups, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta.

@@

Sensitivity to ingroup sampling has received less attention than sensitivity to outgroup sampling. This may be because results have tended to be more sensitive to outgroup sampling.

@@

### Alignment composition

XXX

**Fig XXTaxon_composition.** Each of the primary matrices considered here, color coded by taxon sampling. Horizontal size is proportional to the number of genes (XXOr should it be sites?) sampled, vertical size to the number of taxa sampled.

XXX

**Fig XXTaxon_composition.** Each of the primary matrices considered here, color coded by the types of genes sampled (XX Ribosomal proteins, etc). Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.

@@

XXX

**Fig XXAlignment overlap.** Pairwise overlap between each of the primary matrices considered here. Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled. The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.

@@

One of the challenges of interpreting support for the placement of the animal root across studies is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways. Here we reanalyze the primary matrices from each study under consistent conditions with iqtree. We selected this tool because it has greater model flexibility than other tools and is very fast.

We first tested a variety of models for each matrix, and inferred support under the selected model. We then analyzed every matrix under a panel of standard models, including XXX.

@@

Site heterogeneity in equilibrium frequency has been a major concern in tests of Ctenophora-sister and Porifera-sister. This has been addressed with CAT models. iqtree provides a new family of C models that also address site heterogeneity. Given the extensive computational cost and concerns about overparameterization of CAT models, we compared iqtree C results to CAT results for a subset of matrices to see if they give consistent results. This would be of technical interest because it would reduce the cost of accommodating compositional heterogeneity in future analyses.

@@

Based on the variation across analyses, we constructed new matrices with altered taxon and gene sampling to test specific hypotheses about differences in support.

@@

## Author contributions

@@

# Supplemental Information

@@

The data matrix was constructed using a semi-automated approach. Genes were translated into proteins, promiscuous domains were masked, all gene sequences from all species were compared to each other with blastp, genes were clustered based on this similarity with TribeMCL [@Enright:2002uq], and these clusters were filtered to remove those with poor taxon sampling and high rates of lineage-specific duplications. Gene trees were then constructed, and in clades of sequences all from the same species all but one sequence were removed (these groups are often due to assembly errors). The remaining gene trees with more than one sequence for any taxon were then manually inspected. If strongly supported deep nodes indicative of paralogy were found, the entire gene was discarded. If the duplications for a a small number of taxa were unresolved, all genes from those taxa were excluded. Genes were then realigned and sites were filtered with Gblocks [@Castresana:2000vy], resulting in a 77 taxon matrix. Some taxa in this matrix were quite unstable, which obscured other strongly-supported relationships. Unstable taxa were identified with leaf stability indices [@Thorley:1999kg], as implemented in phyutility [@Smith:2008gb], and removed from the matrix. This resulted in the 64-taxon matrix that is the focus of most of their analyses. Phylogenetic analyses were conducted under the F81+CAT model in phylobayes [@Lartillot:2004dq], and under the WAG model in MrBayes [@Ronquist:2003hx] and RAxML [@Stamatakis:2006wc].

@@

Pick *et al.* [@Pick:2010eb] sought to test whether Ctenophora-sister was an artefact of insufficient taxon sampling. They added new and additional published sequence data to the 64-taxon matrix of Dunn *et al.* [@Dunn:2008ky]. The new taxa included 12 sponges, 1 ctenophore, 5 cnidarians, and *Trichoplax*. They further modified the matrix by removing 2,150 sites that were poorly sampled or aligned. They considered two different sets of outgroups: Choanoflagellata (resulting in Choanimalia) and the same sampling as Dunn *et al.* (resulting in Opisthokonta).

All their analyses were conducted under the F81+CAT+Gamma model in phylobayes [@Lartillot:2004dq], in both a Bayesian framework and with bootstrapping. All analyses have the same ingroup sampling and site removal so it isn't possible to independently assess the impact of these factors. Analyses with Choanimalia sampling recovered Porifera-sister with 72% posterior probability (PP) and 91% bootstrap support (BS). With broader Opisthokonta sampling, support for Porifera-sister is 84% PP. This is an interesting case where increased outgroup sampling leads to increased support for Porifera-sister.

The authors argue that previous results supporting Ctenophora-sister "are artifacts stemming from insufficient taxon sampling and long-branch attraction (LBA)" and that "this hypothesis should be rejected". Although the posterior probabilities supporting Porifera-sister are not strong, they conclude:

@@

They also investigated saturation, and conclude that Dunn *et al.* [@Dunn:2008ky] is more saturated than Philippe *et al.* 2009 [Philippe:2009hh]. Note that the Pick *et al.* [@Pick:2010eb] dataset is not reanalyzed here because partition data are not available, and due to site filtering the partition file from Dunn *et al.* [@Dunn:2008ky] cannot be applied to this matrix.
